Type Refinements for the Working Class

### 2015-12-24 paper.tex
Compare with Allen's supervaluation semantics (#9)

We will demonstrate how to define certain refinements, and prove that they are
types.

@@

>>>>

We will demonstrate how to define certain refinements, and prove that they are
types. The refinements below will differ from the ones defined in Section~

@@

\section{Related Work}

In \cite{allen:2006}, Stuart Allen gave a \emph{supervaluation} semantics for
Nuprl sequents involving atoms (symbols) by quantifying over plenty of choices
for the interpretation of atoms---possible interpretations including strings,
finite sets, etc. One benefit of Allen's indirect treatment is that the logical
theory remains profoundly open-ended, in the sense that it is neither the case
that the naturals may be injected into the atoms, nor is it the case that they
cannot.

\newcommand\AtomModel{\mathfrak{A}}
\newcommand\NuprlSeq[2]{\IMode{#1}\gg\IMode{#2}}
\newcommand\NuprlInterp[2]{\IMode{#1}\models\IMode{#2}}
\newcommand\NuprlInterpSeq[3]{\IMode{#1}\models\NuprlSeq{#2}{#3}}
\newcommand\Strings{\mathbf{String}}

Unlike in our treatment, where the judgments of type theory are intensional
with respect to symbol contexts, Allen takes a more model-theoretic approach
and parameterizes the explanation of type theory over a model $\AtomModel$ for
the abstract atoms.

A Nuprl sequent $\Define{\mathfrak{s}}{H\gg P}$ means ``$P$ is true under
hypotheses $H$'', and it may contain (abstract) atoms in its syntax; we write
$\FS{\mathfrak{s}}$ for the set of free atoms of the sequent $\mathfrak{s}$.
Then, a model $\AtomModel$ consists in a carrier set $\Dom{\AtomModel}$ which
shall interpret the atoms, and a valuation
$\NuprlInterp{\AtomModel}{\mathfrak{s}}$ for any Nuprl sequent $\mathfrak{s}$
which takes its atoms from $\Dom{\AtomModel}$.  A model is called
\emph{discrete} in case its carrier set has decidable equality.  For a Nuprl
sequent $\mathfrak{s}$ with atoms in $\AtomModel$ and an injective function
$\Of{\rho}{\Dom{\AtomModel}\inj\Dom{\AtomModel'}}$, let
$\Rename{\rho}{\mathfrak{s}}$ be the sequent with atoms in $\Dom{\AtomModel'}$
got by replacing each atom $a$ with $\rho(a)$.

Then, a supervaluation is given for Nuprl sequents which take atoms in
$\Strings$, written $\NuprlInterp{*}{\mathfrak{s}}$, which holds if there
exists a $\Member{k}{\Naturals}$ such that for every discrete model
$\AtomModel$ having at least $k$ members, for every injection
$\Of{\rho}{\FS{\mathfrak{s}} \inj \AtomModel}$, we have
$\NuprlInterp{\AtomModel}{\Rename{\rho}{\mathfrak{s}}}$. Under the
supervaluation, we can rule out certain observations about atoms which might
hold in particular models.

For instance, assuming a type $\Naturals$ of natural numbers and $\mathbb{A}$
of atoms, with $\Naturals\inj\mathbb{A}$ the type of embeddings of the naturals
into the atoms, we do have
$\NuprlInterpSeq{\Strings}{\cdot}{\Naturals\inj\mathbb{A}}$, because there are
numerous ways to embed the naturals into the strings; however, because the
embedding fails in every finite model $\AtomModel$, we do not have
$\NuprlInterpSeq{*}{\cdot}{\Naturals\inj\mathbb{A}}$.  Furthermore, we also do
not have $\NuprlInterpSeq{*}{\iota:\Naturals\inj\mathbb{A}}{\bot}$, because this
would fail in any infinite model.

\paragraph{Comparison with the present work}

Our explanation of atoms is not equivalent to the supervaluation semantics.
First, let us introduce a notion analogous to the Nuprl sequent: when
$\RefinesCtx{\cdot}{\Phi}{\Gamma}$ and $\FRefines{\cdot}{\Phi}{\phi}{\tau}$, we
can form the sequent $\NuprlSeq{\Phi}{\phi}$, meaning that there exists an
$\Member{\Upsilon}{\SCtx}$ and
$\Member{M}{\Exprs_\tau(\Upsilon\parallel\Gamma)}$ such that
$\FRMemEq{\Upsilon}{\Phi}{M}{M}{\phi}$. We will write
$\models\NuprlSeq{\Phi}{\phi}$ when $\NuprlSeq{\Phi}{\phi}$ obtains.

\begin{thm}
  \label{thm:no-nat-injection}
  As in the supervaluation semantics, we have
  $\not\models\NuprlSeq{\cdot}{\Naturals\inj\mathbb{A}}$.
\end{thm}
\begin{proof}
  Suppose $\models\NuprlSeq{\cdot}{\Naturals\inj\mathbb{A}}$. Then, there
  exists an $\Member{\Upsilon}{\SCtx}$ and $\Member{M}{\Exprs_\tau(\Upsilon)}$
  such that $\RMemEq{\Upsilon}{M}{M}{\Naturals\inj\mathbb{A}}$; but this
  is impossible because it would require a means of embedding the naturals into
  finite $\Dom{\Upsilon}$.
\end{proof}

\begin{thm}
  Contrary to the supervaluation semantics, we do have
  $\models\NuprlSeq{\iota:\Naturals\inj\mathbb{A}}{\bot}$.
\end{thm}
\begin{proof}
  It suffices to show that for all $\Upsilon$, there is no $\iota$ such that
  $\RMemEq{\Upsilon}{\iota}{\iota}{\Naturals\inj\mathbb{A}}$. However, by the
  definition of $\SCtx$, all $\Member{\Upsilon}{\SCtx}$ are finite, so there
  can be no injection of the naturals into $\Dom{\Upsilon}$.
\end{proof}

The crux of the problem is that all worlds (symbol contexts) are finite in our
semantics, so it is possible to observe that it will never be possible to
inject the naturals into the atoms. One way to make our system more open-ended
is to replace the current semantic universe $\Define{\SemU}{\Classes^\SCtx}$
with an extended universe $\Define{\SemU_\infty}{\Classes^{\SCtx_\infty}}$ with
$\SCtx_\infty$ the set of possibly-infinite symbol contexts and injective,
sort-preserving renamings. Externally, by Theorem~\ref{thm:no-nat-injection} we
could still never exhibit an embedding of the naturals, but internally, we
would be unable to rule it out.
